### 2018-05-23 doc/submission/reviewer-response.md

4. While the discussion emphasizes DNL, it does not allude to how DNL can
mechanistically lead to the metabolic endpoints used in this study. It would be
helpful to indicate that DNL is complicated, and in certain tissues is involved
in the promotion of chronic inflammation leading to decreased insulin
sensitivity and other adverse metabolic events (Wei X et al. Nature 2016;
539:294-298).

    *The discussion on DNL has been clarified in light of the reviewer's comments.*

5. On page 9, line 11 from the bottom, the text omits the actual r values for
the relationships between fatty acids and TG (should be in the 0.8 range for
many) and waist circumference (should be in the 0.3 range for many). More
omissions are made on lines 4 and 5 from the bottom when referring to the data
in Figure 3.

    *We thank the reviewer for noticing that omission. The r values have been
    included in the text.*

@@

# Reviewer 2

1.	It is a frustrating that this is seemingly the third in a series of papers
that is looking at patterns in lipid fractions and association with metabolic
outcomes. Thus, the TG, NEFA and phospholipid fractions all seem to be a 'least
publishable unit'. The authors must integrate the knowledge gained from the
previous analysis of NEFA and phospholipids and discuss some of the seemingly
contradictory findings from the previous studies. If the data were examined,
would the DNL story 'hold up'?

    *We thank the reviewer for these comments. We agree that ideally the results
    of all the pools would have been included together, but given the number of
    fatty acids, the longitudinal visits, the depth and volume of the analyses,
    and the complex biology involved, it simply wasn't possible to contain
    everything in a single journal paper. We have included a brief discussion of
    the results from the other fractions in the Discussion section. The 
    complete and integrated discussion and presentation of the results is found
    in the first author's (LWJ) PhD thesis {{link}}. As to a combined analysis,
    we have presented these results at the European Diabetes Epidemiology Group
    Meeting in April, 2018 {{include DOI}}, and the proposed DNL mechanism
    remained clearly associated and had a strong contribution to association
    with diabetes pathogenesis compared to other fatty acids/compositions.
    A short aside, these currently unpublished findings do not invalidate or 
    undermine the previous work. The combined analysis uses a powerful
    statistical technique (PLS), but it has a number of major limitations,
    not least of which is it can only handle cross-sectional data and can not
    adjust for potential confounders. The present analysis uses PLS in addition
    to the powerful longitudinal technique GEE. {{include a short 
    description expanding on these results here or in the manuscript?}}*

@@

2.	The discussion of GEE and its limitations is well done. However, it seems
that the analysis suggests that total triglycerides are predictive of HOMA2-%S
and the other parameters. Do we need to subfractionate to get insight into risk
for type 2 diabetes?

    *This is excellent question. For a couple of reasons, no we don't need to 
    subfractionate TAG to understand risk. Firstly, we found that the four TGFA
    also highly positively correlated with clinical TAG. Secondly, clinical TAG
    had a much higher magnitude of association compared to any individual
    TGFA. Measuring clinical TAG is enough to assess risk. However, our results
    provide insight into the potential mechanisms regarding how TAG influences risk
    for diabetes, which we show could be due to higher amounts of the four TGFA
    and that higher amounts of these four TGFA alone may strongly contribute to
    the higher clinical TAG.*

### 2018-05-23 doc/submission/reviewer-response.md

1. The authors noted that cluster of four TGFA that correlated negatively with
IS (14:0, 16:0, 14:1n-7, 16:1n-7) are products of DNL, and are therefore likely
reflecting a connection between DNL, CHO intake and diabetes risk. This
conclusion seems a bit overstated. Clearly, a link between this cluster and DNL
is possible, but the signature could be also be reflecting dietary intake,
hepatic steatosis and/or TGFA clearance by tissues such as muscle and adipose
tissue. The authors should include a more balanced discussion of potential
explanations/mechanisms beyond DNL.

    *We thank the reviewer for these comments. We have revised the discussion to 
    include alternative possibilities.*

2. Are there any additional analyses that can be offered to strengthen evidence
the TGFA composition is indeed a biomarker of DNL and/or CHO intake is this
specific cohort? For example, did the PROMISE study include food recall
records?

    *PROMISE did collect food frequency questionnaires, however this was done at
    the 3rd year visit while the fatty acids where measured at the baseline visit.
    So while we could technically use these dietary intake data to help understand the fatty acid data,
    we would likely obtain inaccurate or biased results because of the time separation.
    We therefore decided to not analyze this data.*

@@

3. Also, the discussion should mention and consider that blood samples were
acquired after an overnight fast. Does fasting/fed state impact the correlation
between TGFA composition and DNL?

    *Excellent question. In the fed state, DNL would increase due to conversion
    of excess carbohydrates to fatty acids. This in turn would impact the composition
    of primarily the TGFA. In order to control for this source of variability
    in our cohort we only measured TGFA during fasting. We have included a brief
    description of this fed vs fasted state in the limitations section.*

### 2018-05-23 doc/manuscript.Rmd
todo to done for fed/fasting for TGFA

<!--
TGFA can fluctuate substantially throughout the day and to control for this all
PROMISE participants came for their clinic visits in the morning in a fasted
state. Some of the inherent variability in TGFA fluctuation can be managed by
standardizing the data collection time, though there may still be some
{{residual error? random error?}}.
[@Rhee2011a] fasting TG status better as discriminating between DM case vs
control than 2hr-OGTT TG
-->

TGFA were only quantified at the baseline visit and as such we cannot
investigate whether there are concomitant changes in TGFA and the metabolic
measures over time. However, to optimally use GEE to analyze the data and for
interpretation, we used the model to infer that a given value of TGFA could
predict values of insulin sensitivity or beta-cell function over a 6 year
period. This in our view is a strength of our analysis, as it reduces the chance
of reverse causality given the tight integration of the glucose and fatty acid
metabolism pathways, as well as maximizes the specific usage of the GEE
modeling.

>>>>

<!--
TGFA can fluctuate substantially throughout the day and to control for this all
PROMISE participants came for their clinic visits in the morning in a fasted
state. Some of the inherent variability in TGFA fluctuation can be managed by
standardizing the data collection time, though there may still be some
{{residual error? random error?}}.
-->

Only fasting TGFA were quantified, and only at the baseline visit. TGFA 
composition can fluctuate substantially throughout the day, so in order to 
control for this, PROMISE participants came for the clinic visit in the morning
and fasted. There is some evidence to suggest that fasting TG is better able
to discriminate diabetes cases compared to a post-prandial state [@Rhee2011a].
Because TGFA were only measured at the baseline visit, we cannot
investigate whether there are concomitant changes in TGFA and the metabolic
measures over time. However, to optimally use GEE to analyze the data and for
interpretation, we used the model to infer that a given value of TGFA could
predict values of insulin sensitivity or beta-cell function over a 6 year
period. This in our view is a strength of our analysis, as it reduces the chance
of reverse causality given the tight integration of the glucose and fatty acid
metabolism pathways, as well as maximizes the specific usage of the GEE
modeling.
